### 2023-09-13 doc/_bayesian_inference.qmd
several updates, including some of the comments by Griffin

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique relevant barcodes we care
about (hereafter referred to as mutant barcodes), $M$, and the number of unique
neutral barcodes, $N$, and $T$ is the number of time points where measurements
were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineages barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Following the same logic, matrix
$\underline{\underline{R}}$ can be split into two matrices as
$$
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
$${#eq-R-mat_split}
where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
barcode reads time series for each neutral lineage and
$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
mutant lineages.

Our objective is to compute the joint probability
distribution for all relative fitness values given our data. We can express this 
joint posterior distribution using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. To develop our model, we can ignore the denominator on the
right-hand side of @eq-bayes_obj, since it serves as a normalization constant
that does not depend on our parameter of interest. However, computing this 
normalization constant represents a major challenge we will have to overcome. 
For now, we can simply write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. We can always marginalize the joint
distribution to obtain the distribution we are interested in. To include these
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding mutant and neutral barcode frequencies. With these
nuisance variables in hand, the full inference problem we must solve takes the
form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
the statistical dependencies of our observations and latent variables allows us
to rewrite the joint probability distribution as a product of conditional
distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
    \underline{\underline{R}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
    \underline{\underline{R}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material where we
work through each of these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

@@

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta \mid
\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
distribution. As further explained in `[point to SI section]`, the numerical
integration problem is replaced by an optimization problem of the form
$$
q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
D_{KL}(
    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the 
complicated high-dimensional numerical integration problem is transformed into
a gradient descent over the parameters of the approximate posterior distribution
$\phi$ such that it minimizes the KL divergence with respect to the true 
posterior distribution. Although to compute @eq-vi_objective we require the
posterior distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the 
variational free energy [@buckley2017]---is mathematically equivalent to 
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to `[point to SI section]` for a short primer on variational inference.

>>>>

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique, relevant barcodes we care
about (hereafter referred to as mutant barcodes), $M$, and the number of unique
neutral barcodes, $N$, and $T$ is the number of time points where measurements
were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineage barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Following the same logic, matrix
$\underline{\underline{R}}$ can be split into two matrices as
$$
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
$${#eq-R-mat_split}
where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
barcode reads time series for each neutral lineage and
$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
mutant lineages.

Our objective is to compute the joint probability distribution for all relative
fitness values given our data. We can express this joint posterior distribution
using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. When defining our statistical model, we need not to focus on
the denominator on the right-hand side of @eq-bayes_obj, and write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}
However, when implementing the model computationally, the normalization constant
on the right-hand side of @eq-bayes_obj must be computed via an integral of the
form
$$
\pi(\underline{\underline{R}}) = \int d\underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
$${#eq-bayes_evidence}
also known as a marginalization integral.

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. To include these nuisance
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding mutant and neutral barcode frequencies. With these
nuisance variables in hand, the full inference problem we must solve takes the
form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, as depicted in @fig-01(C),
the statistical dependencies of our observations and latent variables allow us
to factorize the joint distribution into the product of multiple conditional
distributions. To exemplify this, let us focus on the variable
$\underline{\underline{F}}^M$ within the diagram in @fig-01(C). The incoming
arrows from $\underline{\bar{s}}_T$ and $\underline{s}^M$

rewrite the joint probability distribution as a product of conditional
distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M,
    \underline{\underline{R}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N,
    \underline{\underline{R}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each term. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material, where we
work through these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

@@

To overcome this computational limitation, in this work, we rely on the recent
development of the automatic differentiation variational inference algorithm
(ADVI) [@kucukelbir2016]. Briefly, when performing ADVI, our target posterior
distribution $\pi(\theta \mid \underline{\underline{R}})$, where $\theta =
(\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})$, is
replaced by an approximate posterior distribution $q_\phi(\theta \mid
\underline{\underline{R}})$, where $\phi$ fully parametrizes the approximate
distribution. As further explained in `[point to SI section]`, the numerical
integration problem is replaced by an optimization problem of the form
$$
q^*_\phi(\theta \mid \underline{\underline{R}}) = \min _\phi
D_{KL}(
    q_\phi(\theta \mid \underline{\underline{R}}) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
$${#eq-vi_objective}
where $D_{KL}$ is the Kulback-Leibler divergence. In other words, the 
complicated high-dimensional numerical integration problem is transformed into
a gradient descent over the parameters of the approximate posterior distribution
$\phi$ such that it minimizes the KL divergence with respect to the true 
posterior distribution. Although to compute @eq-vi_objective, we require the
posterior distribution we are trying to approximate $\pi(\theta \mid
\underline{\underline{R}})$, it can be shown that maximizing the so-called
evidence lower bound (ELBO) [@kingma2014]---equivalent to minimizing the 
variational free energy [@buckley2017]---is mathematically equivalent to 
performing the optimization prescribed by @eq-vi_objective. We direct the reader
to `[point to SI section]` for a short primer on variational inference.

### 2023-09-13 doc/_experiment.qmd
several updates, including some of the comments by Griffin

## Experimental setup

Our inference pipeline is designed for the type of competition assays presented
in @kinsler2020. In these experiments, an ensemble of genotypes is pooled
together with an unlabeled reference strain that, initially, represents the vast
majority ($\geq 90\%$) of the cells in the culture (see schematic in
@fig-01(A)). Furthermore, a subfraction of labeled genotypes equivalent to
the unlabeled reference strain---hereafter defined as *neutral* lineages---are
spiked in at a relatively high abundance ($\approx 3-5\%$), the rest of the
culture being left for the ensemble of genotypes of interest.

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae*.
@fig-01(B) presents a typical barcode trajectory where the darker barcode
trajectories represent the neutral lineages.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages. (C) Bayesian network defining the statistical dependence
between observations (filled circles) and latent variables (empty circles). The
dotted line highlights the set of variables we are interested in. See
@sec-bayesian_inference for specific definitions of each
variable.](./figs/fig01){#fig-01}

>>>>

## Experimental setup{#sec-experiment}

The present work is designed to analyze time-series data of relative abundance
of multiple microbial lineages uniquely identified by a DNA barcode
[@kinsler2020; @ascensao2023]. In these competition assays, an ensemble of
genotypes is pooled together with an unlabeled reference strain that, initially,
represents the vast majority ($\geq 90\%$) of the cells in the culture (see
schematic in @fig-01(A)). Furthermore, a fraction of labeled genotypes
equivalent to the unlabeled reference strain---hereafter defined as *neutral*
lineages---are spiked in at a relatively high abundance ($\approx 3-5\%$). The
rest of the culture is left for the ensemble of genotypes of interest.

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-7 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae* and to
@ascensao2023 for equivalent assays for *E. coli*. @fig-01(B) presents a typical
barcode trajectory where the darker barcode trajectories represent the so-called
*neutral lineages* that are genetically equivalent to the untagged ancestor
strain that initially dominates the culture. These neutral lineages simplify the
inference problem since the fitness of all relevant barcodes is quantified with
respect to these barcodes.

![**Typical competitive fitness experiment**. (A) Schematic of the typical
experimental design to determine the competitive fitness of an ensemble of
barcoded genotypes. Genotypes are pooled together and grown over multiple
growth-dilution cycles. At the end of each cycle, a sample is processed to
generate a library for amplicon sequencing. (B) Typical barcode trajectory
dataset. From each time point, the relative frequency of each barcode is
determined from the total number of reads. Different shades of blue represent
different relative fitness. Darker gray lines define the typical trajectory of
neutral lineages. (C) Bayesian network defining the statistical dependence
between observations (filled circles) and latent variables (empty circles). The
dotted circle highlights the set of variables we are interested in. One-way
arrows depict the statistical dependence between variables. For example, the
value of $\underline{\underline{F}}^N$ only depends on the value of
$\underline{\bar{s}}_T$, while $\underline{\underline{F}}^M$ depends on
$\underline{\bar{s}}_T$ and $\underline{s}^M$. See @sec-bayesian_inference for
specific definitions of each variable.](./figs/fig01){#fig-01}

### 2023-09-13 doc/_fitness_model.qmd
several updates, including some of the comments by Griffin

Empirically, the barcode relative frequency trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the reference strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
@eq-fitness as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logfreq}
@eq-logfreq separates the observables---the barcode frequencies---from the
latent parameters we want to infer from the data---the population mean fitness
and the barcode relative fitness. This is ultimately the functional form used in
our inference pipeline. Therefore, the relative fitness is computed by knowing
the log frequency ratio of each barcode throughout the growth-dilution cycles.

It is important to notice that the frequencies $f_{t}^{(b)}$ are not the allele
frequencies in the population (most of the culture is not sequenced since the
reference strain is not barcoded), but rather the relative frequencies in the
total number of sequencing reads. A way to conceptualize this subtle but
important point is to assume exponential growth in the *number of cells*
$N_t^{(b)}$ of the form
$$
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
$${#eq-ncells}
for every barcode $b$ with growth rate $\lambda^{(b)}$. However, when we
sequence barcodes, we do not directly measure the number of cells, but some
number of reads $r_t^{(b)}$ that map to barcode $b$. In the simplest possible
scenario, we assume
$$
r_{t}^{(b)} \propto N_{t}^{(b)},
$${#eq-r_to_n}
where, importantly, the proportionality constant depends on the total number of
reads for the time $t$ library, which might vary from library to library.
Therefore, to compare the number of reads between libraries at different time
points, we must normalize the number of reads to the same scale. The simplest
form is to define a relative abundance, i.e., a frequency with respect to the
total number of reads,
$$
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
$${#eq-r_to_n_norm}
This is the frequency @eq-fitness describes.

>>>>

Empirically, the barcode relative frequency trajectories follow an exponential
function of the form [@levy2015; @kinsler2020; @ascensao2023]
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the reference strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
@eq-fitness as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logfreq}
@eq-logfreq separates the measurements---the barcode frequencies---from the
unobserved---sometimes referred to as latent---parameters we want to infer from
the data---the population mean fitness and the barcode relative fitness. This is
ultimately the functional form used in our inference pipeline. Therefore, the
relative fitness is computed by knowing the log frequency ratio of each barcode
throughout the growth-dilution cycles.

It is important to notice that the frequencies $f_{t}^{(b)}$ are not the allele
frequencies in the population (most of the culture is not sequenced since the
reference strain is not barcoded), but rather the relative frequencies in the
total number of sequencing reads. A way to conceptualize this subtle but
important point is to assume exponential growth in the *number of cells*
$N_t^{(b)}$ of the form
$$
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
$${#eq-ncells}
for every barcode $b$ with growth rate $\lambda^{(b)}$. However, when we
sequence barcodes, we do not directly measure the number of cells, but some
number of reads $r_t^{(b)}$ that map to barcode $b$. In the simplest possible
scenario, we assume
$$
r_{t}^{(b)} \propto N_{t}^{(b)},
$${#eq-r_to_n}
where, importantly, the proportionality constant depends on the total number of
reads for the library at time $t$, which might vary from library to library.
Therefore, to compare the number of reads between libraries at different time
points, we must normalize the number of reads to the same scale. The simplest
form is to define a relative abundance, i.e., a frequency with respect to the
total number of reads,
$$
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
$${#eq-r_to_n_norm}
This is the frequency @eq-fitness describes.

### 2023-09-13 doc/_hierarchical_replicates_results.qmd
several updates, including some of the comments by Griffin

## Accounting for experimental replicates via hierarchical models

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as completely independent, losing the power to
utilize multiple measurements when trying to learn a single parameter. This can
hurt the inference since, in principle, the value of our parameter of interest
should not depend on the particular experimental replicate in question. However,
this approach does not allow us to properly "combine" the uncertainties in both
experiments when performing the inference. On the other hand, we can pool all
data together and treat our different experiments as a single measurement with
higher coverage. This loses the subtle differences that come from biotic and
abiotic batch effects, effectively halving the data that goes into our inference
problem.

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the ultimate parameter we are interested in---for our case we have a
hyper-fitness value $\theta^{(m)}$ for each mutant. Second, each experiment
draws randomly from the distribution of this hyper-parameter, allowing for
subtle variability between experiments to be accounted for---for our inference
pipeline, each experimental replicate gets assigned a fitness value $s^{(m,i)}$,
where the extra index indicates the $i$-th experimental replicate. As
schematized by the Bayesian network on @fig-03(A) and further detailed in
`[point to SI section]`, our statistical model can easily be extended to a
hierarchical structure, where the fitness of each relevant barcode on each
experimental replicate is a sample from a global hyper-fitness parameter.

To test the performance of this model, we simulated two experimental replicates
with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
truth hyper-fitness value $\theta^{(m)}$ for each barcode. To capture
experimental batch effects, we sampled a variation from this hyper-fitness value
for each experimental replicate $s^{(m, i)}$. @fig-03(D) shows the relationship
between hyper-fitness and replicate fitness values for this simulation. As shown
by the posterior predictive checks examples in @fig-03(E), the hierarchical
model can correctly fit the data for each experimental replicate. Furthermore,
@fig-03(F-G) show a high degree of correlation between the ground truth and the
inferred fitness values. The ECDFs shown in @fig-03(H-I) reveal that for
$\approx 75\%$ of the non-neutral barcodes, the ground truth hyper-fitness
values fall within one standard deviation from the mean value in the posterior
distributions.

![**Hierarchical model on experimental replicates**. (A) Bayesian network
schematizing the statistical dependence between observations (filled circles)
and latent variables (empty circles). The replicates are connected via the
hyper-parameters $\underline{\theta}^M$. See `point to SI` for detailed
definitions of each variable. (B-C) Simulated replicate datasets with 900
barcodes of interest and 100 neutral lineages. (D) Comparison between the
simulation ground truth hyper-fitness and each replicate ground truth fitness.
The scatter between parameters captures experimental batch effects. (E) Examples
of the posterior predictive checks for all neutral lineages (upper left panels)
and a subset of representative mutant lineages. Shaded regions from light to
dark represent the 95\%, 68\% and 5\% credible regions. (F-G) Comparison between
the ground truth hyperfitness (F) and replicate fitness (G) values from the
simulation and the inferred parameters. Gray error bars represent the 68\%
posterior credible region. (H-I) The empirical cumulative distribution function
(ECDF) for the absolute z-score value of the ground truth parameter value within
the inferred hyper-fitness posterior distribution (H) and replicate fitness
(I).](./figs/fig03){#fig-03}

As shown in @fig-04, the structure imposed by the hierarchical model, where we
explicitly account for the connection between experimental replicates can
improve the quality of the inference.  Inferred fitness values between
experimental replicate exhibit a stronger degree of correlation in the
hierarchical model (@fig-04(A)) compared to conducting inference on each
replicate independently (@fig-04(B)). Moreover, when comparing the inferred
hyper-fitness values---the objective parameter when performing multiple
experimental measurements---the hierarchical model outperforms averaging the
independent experimental replicates as shown in @fig-04(C) and (D).

![**Comparison between hierarchical model and single dataset model**. (A-B)
comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model (A) or independently fitting each dataset (B). Gray
error bars represent the 68\% credible regions. (C) Comparison between the
ground truth hyper-fitness value and the inferred parameters. The blue dots show
the inferred hyper-fitness values when assuming a hierarchical model. Gray error
bars show the 68\% credible region for this inference. The yellow dots show the
average of the mean inferred fitness values for the two experimental replicates.
No error bars are shown for these as it is inappropriate to compute one with two
data points. (D) Empirical cumulative distribution function (ECDF) of the
absolute difference between the inferred mean and the ground truth
hyper-fitness.](./figs/fig04){#fig-04}

>>>>

## Accounting for experimental replicates via hierarchical models{#sec-replicates}

Our inference pipeline can be extended to account for multiple experimental
replicates via Bayesian hierarchical models [@betancourt2013]. Briefly, when
accounting for multiple repeated measurements of the same phenomena, there are
two extreme cases one can use to perform the data analysis: On the one hand, we
can treat each measurement as entirely independent, losing the power to utilize
multiple measurements when trying to learn a single parameter. This can hurt the
inference since, in principle, the value of our parameter of interest should not
depend on the particular experimental replicate in question. However, this
approach does not allow us to properly "combine" the uncertainties in both
experiments when performing the inference. On the other hand, we can pool all
data together and treat our different experiments as a single measurement with
higher coverage. This loses the subtle differences from biotic and abiotic batch
effects, effectively halving the data that goes into our inference problem.

Hierarchical models present a middle ground between these extremes. First,
hierarchical models rely on the definition of so-called *hyper-parameters*, that
capture the parametric inference we are interested in---for this inference
problem, we have a hyper-fitness value $\theta^{(m)}$ for each mutant. Second,
each experiment draws randomly from the distribution of this hyper-parameter,
allowing for subtle variability between experiments to be accounted for---in the
present inference pipeline, each experimental replicate gets assigned a *local*
fitness value $s^{(m)}_i$, where the extra sub-index indicates the $i$-th
experimental replicate. For our hierarchical model, these local fitness values
interact via the common dependency on the global hyper-fitness value. As
schematized in @fig-03(A), this interaction runs both ways. First, the data from
one experimental replicate ($\underline{\underline{R}}^M_k$ in the diagram)
informs all local fitness values via the global hyper-fitness (upper panel in
@fig-03(A)). Second, the local fitness value is informed by the data from all
experimental replicates via the same global hyper-fitness parameter (lower panel
in @fig-03(A)).

To test the performance of this model, we simulated two experimental replicates
with 1000 unique barcodes (see @fig-03(B-C)) where we randomly sampled a ground
truth hyper-fitness value $\theta^{(m)}$ for each barcode. We sampled a
variation from this hyper-fitness value for each experimental replicate $s^{(m,
i)}$ to capture experimental batch effects. @fig-03(D) shows the relationship
between hyper-fitness and replicate fitness values for this simulation. The
posterior predictive checks examples in @fig-03(E) show that the hierarchical
model can correctly fit the data for each experimental replicate. Furthermore,
@fig-03(F-G) show a high correlation between the ground truth and the inferred
fitness values. The ECDFs shown in @fig-03(H-I) reveal that for $\approx 75\%$
of the non-neutral barcodes, the ground truth hyper-fitness values fall within
one standard deviation from the mean value in the posterior distributions.

![**Hierarchical model on experimental replicates**. (A) Schematic depiction of
the interactions between local fitness values $\underline{s}_l^M$ through the
global hyper-fitness value $\underline{\theta}^M$ for $K$ hypothetical
experimental replicates. The upper diagram shows how the data from replicate $k$
informs all local fitness values via the hyper-fitness parameter. The lower
panel shows the reverse, where all datasets inform the local fitness value.
(B-C) Simulated replicate datasets with 900 barcodes of interest and 100 neutral
lineages. (D) Comparison between the simulation ground truth hyper-fitness and
each replicate ground truth fitness. The scatter between parameters captures
experimental batch effects. (E) Examples of the posterior predictive checks for
all neutral lineages (upper left panels) and a subset of representative mutant
lineages. Shaded regions from light to dark represent the 95\%, 68\%, and 5\%
credible regions. (F-G) Comparison between the simulation's ground truth
hyper-fitness (F) and replicate fitness (G) values and the inferred parameters.
Gray error bars represent the 68\% posterior credible region. (H-I) The
empirical cumulative distribution function (ECDF) for the absolute z-score value
of the ground truth parameter value within the inferred hyper-fitness posterior
distribution (H) and replicate fitness (I).](./figs/fig03){#fig-03}

As shown in @fig-04, the structure imposed by the hierarchical model schematized
in @fig-03(A), where we explicitly account for the connection between
experimental replicates can improve the quality of the inference. Inferred
fitness values between experimental replicates exhibit a stronger degree of
correlation in the hierarchical model (@fig-04(A)) compared to conducting
inference on each replicate independently (@fig-04(B)). Moreover, when comparing
the inferred hyper-fitness values---the objective parameter when performing
multiple experimental measurements---the hierarchical model outperforms
averaging the independent experimental replicates as shown in @fig-04(C) and
(D).

![**Comparison between hierarchical model and single dataset model**. (A-B)
comparison of inferred fitness values between experimental replicates when
fitting a hierarchical model (A) or independently fitting each dataset (B). Gray
error bars represent the 68\% credible regions. (C) Comparison between the
ground truth hyper-fitness value and the inferred parameters. The blue dots show
the inferred hyper-fitness values when assuming a hierarchical model. Gray error
bars show the 68\% credible region for this inference. The yellow dots show the
average of the mean inferred fitness values for the two experimental replicates.
No error bars are shown for these, as it is inappropriate to compute one with
two data points. (D) Empirical cumulative distribution function (ECDF) of the
absolute difference between the inferred mean and the ground truth
hyper-fitness.](./figs/fig04){#fig-04}
